---
title: "Lemmings data cleaning"
format: html
---



```{r}
#| message: false  

# Libraries
library(tidyverse)
library(here)
library(naniar)

# file names
datadir_raw <- "01_data/data-raw/"

datadir_processed <- "01_data/data-processed/"

lemmings_file <- "ASDN_Daily_pred_lemm.csv"

observers_file <- "observers.csv"

```


## Lemmings survey data

### Data Exploration

Import the data

```{r}
# Import the species
lemmings_data <- read_csv(here(datadir_raw, lemmings_file))

glimpse(lemmings_data)

```

All the fields seem to have a correct data type, except for Date, suggesting that no unexpected value types are present. Some field are categorical and we will check the unique values for those and compare those to the existing metadata.

Check NAs frequency

```{r}
lemmings_data %>% 
   miss_var_summary()
```

Notes have mostly missing values, which is expected for such open ended field. 11% of missing data for Observer is somewhat more problematic, but will not perclude analysis. The Location values are not matching the other tables and will be dropped.


Let's have a look at the Notes:

```{r}
lemmings_data %>% 
  distinct(Notes)
```

Might be useful information in general but not for our teaching purposes, so we will drop it as well.

Check Categorical fields

```{r}
lemmings_data %>% 
  distinct(Location)
```

It seems that the Location field is not well standardized and might not add much information. We will drop that attribute.


### Data Cleaning

#### Observers

```{r}
lemmings_data %>% 
  distinct(Observer)
```

Load the Observers from the table that was derived from the Bird nest table

```{r}
observers_nests <- read_csv(here(datadir_processed, observers_file))
```
Let's see the observers that are not listed in the bird nest observation table

```{r}
anti_join(lemmings_data, observers_nests, join_by(Observer == Abbreviation)) %>%
  distinct(Observer)
```

We have many occurrence of multiple observers listed in one cell. We will take the same approach as for the bird nests table and arbitrarily keep only the first observer.  However for now, let's split those multiple entries into separate row and produce a list of observers who do not yet exists in our observer table derived from the nest observation:

```{r}
missing_observers <- lemmings_data %>%
  separate_rows(Observer, sep="[,+/]") %>%  # split multiple entries per cell into separate rows
  mutate(Observer = trimws(Observer)) %>%  # some observers have leading spaces
  distinct(Observer) %>%
  full_join(observers_nests, join_by(Observer == Abbreviation), keep = TRUE) %>% 
  arrange(Observer) %>%
  filter(is.na(Abbreviation))
```

Export the csv to edit manually (need to invent names)

```{r}
write_csv(missing_observers, here(datadir_processed, "lemming_observers_missing.csv"))
```


Manually editing this file in Excel:

- inventing names that matched the initial + last name logic for the abreviation
- dropped rows containing words like: all, crew, weather, ...


```{r}
observers_fixed <- read_csv(here(datadir_processed, "lemming_observers_fixed.csv"))
```

Merge with the exisintg oberser table:

```{r}
observers_all <- bind_rows(observers_nests, observers_fixed)
```


Keep only the first observer for multiple observers entries and remove any spaces

```{r}
lemmings_first_obs <- lemmings_data %>%
  mutate(Observer = trimws(str_split_i(Observer, pattern = "[,+/]", 1))) %>%
  mutate(Observer = ifelse(Observer == "kshiemrief", "kschiemrief", Observer)) # fix misspelling
```

Check which Observer values do not exist in the observer_all table

```{r}
bad_observers <- lemmings_first_obs %>%
anti_join(observers_all, join_by(Observer == Abbreviation)) %>%
  distinct(Observer)

bad_observers
```

Those are all bad entries for Observers, we can set them to NA

```{r}
lemmings_good_obs <- lemmings_first_obs %>%
  mutate(Observer = ifelse(Observer %in% pull(bad_observers), NA, Observer)) 
```

Update the observers table with all the good observers of lemmings & co

```{r}
observers_all %>%
  filter(!Abbreviation %in% bad_observers) %>%
  arrange(Abbreviation) %>%
  write_csv(here(datadir_processed, "observers_all.csv"))
```


#### Species

```{r}
lemmings_good_obs %>%
  distinct(Species)
```

That's a lot of unique species... probaly more any predators than just lemmings


Let's see if those sepcies are all in our species table"

```{r}
species_data <- read_csv(here(datadir_processed, "species.csv"))

lemmings_good_obs %>%
anti_join(species_data, join_by(Species == Code)) %>%
  count(Species)

```

15 species are not in our master species table. However they only account for 53 row in our table, so we will simply drop those rows:

```{r}
lemmings_good_sp <- lemmings_good_obs %>% 
  filter(Species %in% species_data$Code)
```


#### Other categorical attributes

Team count should be y or n, but we have a few N. Assuming it is also no:

```{r}
lemmings_good_sp %>% count(Count_type)
```

Looks good!

```{r}
lemmings_good_sp %>% count(Team_count)
```

Team_count should be y or n, but we have a few N. Assuming it is also no:

```{r}
lemmings_good_sp <- lemmings_good_sp %>%
  mutate(Team_count = tolower(Team_count))
```


#### Dates

Let's see if we get errors when trying to convert to the date format:

```{r}
tt <- lemmings_good_sp %>% 
  mutate(Date2 = dmy(Date),
         Year2 = year(Date2),
         Year_comp = ifelse(Year == Year2, TRUE, FALSE))
```

```{r}
tt %>% filter(Year_comp == FALSE) %>% select(Year, Date, Date2, Year2, Year_comp)
tt %>% filter(is.na(Date2)) %>% select(Year, Date, Date2, Year2, Year_comp)
```

No errors, although `as_date()` would have guessed the wrong format. Let's implement the change:

```{r}
lemmings_good_date <- lemmings_good_sp %>% 
  mutate(Date = dmy(Date))
```


Let's write the cleaned Lemmings table to a csv file:

```{r}
lemmings_good_date %>%
  select(-Location, -Notes) %>%    # no valuable info in those columns
  filter(!is.na(Start_time) ) %>%  # will have to be part of the primary key (loosing 44 entries)
  write_csv(here(datadir_processed, "lemmings_fixed.csv"))
```

