---
title: "Lemmings data cleaning"
format: html
---



```{r}
#| message: false  

# Libraries
library(tidyverse)
library(here)
library(naniar)

# file names
datadir_raw <- "01_data/data-raw/"

datadir_processed <- "01_data/data-processed/"

lemmings_file <- "ASDN_Daily_pred_lemm.csv"

observers_file <- "observers.csv"

```


## Lemmings survey data

### Data Exploration

Import the data

```{r}
# Import the species
lemmings_data <- read_csv(here(datadir_raw, lemmings_file))

glimpse(lemmings_data)

```

All the fields seem to have a correct data type, suggesting that no unexpected value types are present. Some field are categorical and we will check the unique values for those and compare those to the existing metadata.


Check NAs frequency

```{r}
lemmings_data %>% 
   miss_var_summary()
```

Notes have mostly missing values, which is expected for such open ended field. 11% of missing data for Observer is somewhat more problematic, but will not perclude analysis. The missing values in the Location files are more problematic since it needs to be used when joining other tables.


Let's have a look at the Notes:

```{r}
lemmings_data %>% 
  distinct(Notes)
```

Might be useful information, so we will keep it desptite this field being sparse.

Check Categorical fields

```{r}
lemmings_data %>% 
  distinct(Location)
```

It seems that the Location field is not well standardized and might not add much information. We will drop that attribute.

```{r}
lemmings_data %>% 
  distinct(Observer )
```


Load the Observers from the table that was derived from the Bird nest table

```{r}
observers_data <- read_csv(here(datadir_processed, observers_file))
```
Let's see the observers that are not listed in the bird nest observation table

```{r}
anti_join(lemmings_data,observers_data, join_by(Observer == Abbreviation)) %>%
  distinct(Observer)
```

We have many occurrence of multiple observers listed in one cell. We will take the same approach as for the bird nests table and arbitrarily keep only the first observer.  However for now, let's split those multiple entries into separate row and produce a list of observers who do not yet exists in our observer table derived from the nest observation:

```{r}
missing_observers <- lemmings_data %>%
  separate_rows(Observer, sep="[,+/]") %>% # split multiple entries per cell into separate rows
  distinct(Observer) %>%
  full_join(observers_data, join_by(Observer == Abbreviation), keep = TRUE) %>% 
  arrange(Observer) %>%
  filter(is.na(Abbreviation))
```

Export the csv to edit manually (need to invent names)

```{r}
write_csv(missing_observers, here(datadir_processed, "lemming_observers_missing.csv"))
```


Manually editing this file in Excel:

- inventing names that matched the initial + last name logic for the abreviation
- dropped rows containing words like all, crew, weather, ...


```{r}
observers_fixed <- read_csv(here(datadir_processed, "lemming_observers_fixed.csv"))
```

