---
title: "Observers"
format: html
---

```{r}
#| message: false  

# Libraries
library(tidyverse)
library(here)


# file names
datadir_raw <- "01_data/data-raw/"

datadir_processed <- "01_data/data-processed/"

lemmings_file <- "ASDN_Daily_pred_lemm.csv"

observers_file <- "observers.csv"

nests_file <- "bird_nests.csv"

```

Load the Observers from the table that was derived from the cleaned camp assignment table (see main_data-cleaning.ipynb)

```{r}
observers_camp <- read_csv(here(datadir_processed, observers_file))
```


## Lemmings Observers

Import the data

```{r}
# Import the species
lemmings_data <- read_csv(here(datadir_raw, lemmings_file))
```

What are the distincts observers:

```{r}
lemmings_data %>% 
  distinct(Observer)
```

There are sometimes many observers listed, let's split this:

```{r}
lemmings_obs <- lemmings_data %>%
  separate_rows(Observer, sep="[,+/]") %>%  # split multiple entries per cell into separate rows
  mutate(Observer = trimws(Observer)) %>%  # some observers have leading spaces
  distinct(Observer)
```


Let's see the observers that are not listed in the camp assignment table

```{r}
anti_join(lemmings_obs, observers_camp, join_by(Observer == Abbreviation)) %>%
  distinct(Observer)
```

Let's produce a list of observers who do not yet exists in our observer table derived from the camp assignment:

```{r}
missing_observers <- lemmings_obs %>%
  full_join(observers_camp, join_by(Observer == Abbreviation), keep = TRUE) %>% 
  arrange(Observer) %>%
  filter(is.na(Abbreviation)) 
```

Export the csv to edit manually (need to invent names)

```{r}
write_csv(missing_observers, here(datadir_processed, "lemming_observers_missing.csv"))
```


Manually editing this file in Excel:

- inventing names that matched the initial + last name logic for the abreviation
- dropped rows containing words like: all, crew, weather, ...


Import the cleaned observer list

```{r}
observers_predfixed <- read_csv(here(datadir_processed, "lemming_observers_fixed.csv"))
```

Merge with the existing camp observer table:

```{r}
observers_camppred <- bind_rows(observers_camp, observers_predfixed)
```


## Nests Observers

```{r}
# Import the species
nests_data <- read_csv(here(datadir_processed, nests_file))
```

What are the distincts observers:

```{r}
nests_data %>% 
  distinct(Observer)
```

Let's see the observers that are not listed in the camp assignment  table

```{r}
anti_join(nests_data, observers_camppred, join_by(Observer == Abbreviation)) %>%
  distinct(Observer) %>% 
  arrange(Observer)
```

Some are clearly breaking the rule of the First Name initial + last name, however it is still quite a long list!

Will export and edit by hand:

- remove any abbreviation <=3 and all capitalized (probably initials)
- anything with crew, all, ....
- Invent names for the renaming entries

Export the csv to edit manually (need to invent names)

```{r}
missing_nestobs <- nests_data %>%
  distinct(Observer) %>%
  full_join(observers_camppred, join_by(Observer == Abbreviation)) %>% 
  arrange(Observer) %>%
  filter(is.na(Name)) %>%
  rename(Abbreviation = "Observer")
```

```{r}
write_csv(missing_nestobs, here(datadir_processed, "nest_observers_missing.csv"))
```


Import the cleaned observer list:

```{r}
observers_nestfixed <- read_csv(here(datadir_processed, "nest_observers_fixed.csv"))
```


Merge with the existing camp observer table:

```{r}
observers_all <- bind_rows(observers_camppred, observers_nestfixed)
```

Check for duplicates

```{r}
observers_all %>%
  group_by(Abbreviation) %>% 
  filter(n()>1) %>% 
  arrange(Abbreviation)
```

Let's keep the first one:

```{r}
observers_nodup <- observers_all %>%
  group_by(Abbreviation) %>% 
  slice(1) %>%
  ungroup()
```


Write the observers table with all the good observers of lemmings & nests:

```{r}
observers_nodup %>%
  arrange(Abbreviation) %>%
  write_csv(here(datadir_processed, "observers_all.csv"))
```


## Update the bird nest data to drop rows with bad observer abbreviations

There were two cases with a period separating the first initial and the last name. It was decided when amnually editing nest observers to update them removing the period to match the abbreviation rule:

```{r}
nests_data <- nests_data %>%
  mutate(Observer = replace_values(Observer, 
                                  "H.Specht" ~ "hspecht",
                                  "d.taylor" ~ "dtaylor",
                                  "eaba_crew" ~ "mchown"    # arbitrarily pick an observer that was at this site at this time period to not loose 112 rows
                 )
         )
```


```{r}
nests_data_obsname <- left_join(nests_data, observers_nodup, join_by(Observer == Abbreviation)) %>%
  filter(!is.na(Name))
```

Loosing about 2000 rows :/


```{r}
nests_data_obsname %>%
  select(-Name) %>%   # not needed
  write_csv(here(datadir_processed, "bird_nests_abbrev.csv"))
```



Since we lost rows in the nests table, we might have lost some nest_id as well. We need to check again that there are no nest_ids in the eggs table that are not in the nest table.


Le
```{r}

(unique(nests_data$Nest_ID) %in% unique(nests_data_obsname$Nest_ID))
```

```{r}
eggs_data <-  read_csv(here(datadir_processed, "bird_eggs.csv"))
```




